---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)
sce <- DropletUtils::read10xCounts(c(pbmc5k="/t1-data/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/raw_feature_bc_matrix")) #(c(pbmc5k)= to create a name vector to replace the location
  

sce
sce[1:5,1:5]
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}

sce <- sce[, colSums(counts(sce)) >0] #sce <- sce[, colSums(counts(sce)) !=0] counts is an assay that is in the 10x, very confusing

sce


```

- Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)
?DropletUtils::emptyDrops()
out <- DropletUtils::emptyDrops(counts(sce), lower=100)

# NA is the UMIs below 100, they are not calculated but they are included
# FDR p-value
out
```

**How do you read the output? Which droplets would you consider empty?** NA

> Answer:
> 
>
>
> 

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}
is.cell <- out$FDR <= 0.01
sum(is.cell, na.rm=TRUE)
sce_filtered_emptydrops <- sce[,which(is.cell), drop=FALSE] #which will remove FALSES and NAs
dim(sce_filtered_emptydrops)
 dim(is.cell)
# Check if p-values are lower-bounded by 'niters'
# (increase 'niters' if any Limited==TRUE and Sig==FALSE)
table(Sig=is.cell, Limited=out$Limited)
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}


plot_data <- tibble(sce_filtered_emptydrops
  )

ggplot(sce_filtered_emptydrops, aes()   ) +
  
  
  
```

```{r}
is.mt <- 
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- read10xCounts(c(pbmc_filtered="/t1-data/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/filtered_feature_bc_matrix/"))
  
  
colData(sce_filtered_cellranger)



all_barcodes <- sce$Barcode

length(all_barcodes)

compare_table <- tibble(
  emptydrops = all_barcodes %in% sce_filtered_emptydrops$Barcode,
  cellranger = all_barcodes %in% sce_filtered_cellranger$Barcode #cellranger was uploaded from the file
)
table(compare_table) #comparison between the filtering of emptydrops and cell ranger. Overall good agreement. If there is not agreement, additional filtering is necessary
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.
Should be done after excluding empty droplets

- Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(1)

sce_scdblfinder <- scDblFinder::scDblFinder(sce_filtered_emptydrops)

colData(sce_scdblfinder) 

table(sce_scdblfinder$scDblFinder.class)

hist(sce_scdblfinder$scDblFinder.score, breaks = 100, xlab = "Double finder score") #higher score, higher the doblets
  
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer:
> 
> 
> 
> 

- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.

```{r}

sce_filtered_scdblfinder <- sce_scdblfinder[,sce_scdblfinder$scDblFinder.class=="singlet"]

table(sce_filtered_scdblfinder$scDblFinder.class)

colData(sce_filtered_scdblfinder)

sce_scdblfinder <- addPerCellQC(sce_scdblfinder)

colData(sce_scdblfinder)
```
```{r}
sce_scdblfinder

ggplot(as.data.frame(colData(sce_scdblfinder))) + 
  geom_violin(aes(x= sce_scdblfinder$scDblFinder.class, y=sum)) +
         labs(x= "droplets info", y= "library size") +
          theme_classic() +
        geom_jitter(aes(x= sce_scdblfinder$scDblFinder.class, y=sum), size=0.5)
  
```


## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
?scuttle::addPerCellQC()
sce <- scuttle::addPerCellQC(sce)
sce

sce$is.cell <- is.cell
length(is.cell)
colData(sce)

ggplot(as.data.frame(colData(sce))) +
         geom_violin(aes(x= is.cell, y=sum), scale = "width") +
         labs(x= "droplets info", y= "library size") +
         coord_cartesian(ylim=c(0,20000)) +
          theme_classic() +
        geom_jitter(aes(x= is.cell, y=sum), size=0.5)
  
 #NA barcodes with less than 200 UMIs
# FALSE has been tested and is an empty droplet 
  
```

```{r}


is.mit <- grep("^MT-", rowData(sce_scdblfinder)$Symbol)

sce_scdblfinder <- scuttle::addPerCellQC(sce_scdblfinder, subsets = list("is.mit" = is.mit))

colData(sce_scdblfinder)
        
mt_plot <- ggplot(as.data.frame(colData(sce_scdblfinder))) +
                  geom_violin(aes(x=scDblFinder.class, y= subsets_is.mit_percent))+
  labs(x= "Singlet or Doublet", y="Percentage mitochrondrial") +
  geom_jitter(aes(x=scDblFinder.class, y= subsets_is.mit_percent), size=0.5); mt_plot
  
  
  
```

```{r}

  
```




- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}

set.seed(1)
sce_decontX <- celda::decontX(sce_filtered_scdblfinder,
                              z=NULL,
                              batch=NULL,
                              maxIter=500)

colData(sce_decontX)
metadata(sce_decontX)
str(metadata(sce_decontX))
```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
set.seed(1)
colData(sce_decontX)
?plotDimReduceCluster
reducedDimNames(sce_decontX)

umap <- reducedDim(sce_decontX, "decontX_UMAP")
umap
plotDimReduceCluster(
                      sce_decontX$decontX_clusters, 
                      dim1=  umap[,1],
                      dim2= umap[,2]
                     )
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
?plotDecontXContamination
plotDecontXContamination(sce_decontX,
  batch = NULL,
  colorScale = c("blue", "green", "yellow", "orange", "red"),
  size = 1   )
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> 
> 

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}

sce_decontX <- logNormCounts(sce_decontX, assay.type = "counts")
rownames(sce_decontX) <- uniquifyFeatureNames(
  ID = rownames(sce_decontX),
  names = rowData(sce_decontX)[["Symbol"]]
)

?uniquifyFeatureNames
assayNames(sce_decontX)
listofgenes <- c("CD3D","CD3E","GNLY","LYZ","S100A8","S100A9","CD79A","CD79B","MS4A1")
plotDimReduceFeature(x= assay(sce_decontX, "logcounts"),
                     features= listofgenes, 
                     dim1=  umap[,1],
                      dim2= umap[,2])

rownames(sce_decontX) <- uniquifyFeatureNames(
)
```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}
markers <- list(
  Tcell_Markers = c("CD3E", "CD3D"),
  Bcell_Markers = c("CD79A", "CD79B", "MS4A1"),
  Monocyte_Markers = c("S100A8", "S100A9", "LYZ"),
  NKcell_Markers = "GNLY"
)
cellTypeMappings <- list(
  Bcells = 4,
  Monocytes = 2,
  Tcells = c(1, 3),
  NKcells = c(5, 6)
)
plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("counts", "decontXcounts")
)

```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  
  
  
  
)
```

- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r}
plotDecontXMarkerExpression(
  
  
  
  
)
```

