---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?

Doesn't show anything because no information was added on how it should be displayed

```{r}
ggplot(diamonds)

```

> Answer:

Doesn't show the table

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?



```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:

It shows the price and carat x and y axis, but doesn't show any values. But it shows a range of values in x and y axis.

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.5)
  
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour=cut))
  
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(diamonds, aes(x = depth, y = price)) +
geom_point(size = 0.7, aes(colour=carat)) +
geom_smooth() 


```
```{r}
head(diamonds)


```

- Assign that last plot to an object called `obds_diamonds`.

```{r}
obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour=cut)) +
geom_smooth() 


```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price, colour=cut)) +
  geom_point(size = 0.7) +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:

# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
head(ChickWeight)
ggplot(ChickWeight, aes(x = Time, y = weight)) +
       geom_point(aes(colour=Diet)) 

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}

head(ChickWeight)
ggplot(ChickWeight, aes(x = Time, y = weight)) +
       geom_point(aes(colour=Diet)) 
```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(ChickWeight, aes(x = Time, y = weight, colour=Diet)) +
       geom_point() +
geom_smooth(method = lm)

```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}
colnames(ChickWeight)
ggplot(ChickWeight, aes(x = Time, y = weight, colour=Diet)) +
       geom_point() +
  geom_smooth(method =lm)+
facet_wrap(~Diet, nrow=1)


```

- Assign that last plot to an object called `obds_chickweight`.

```{r}

obds_chickweight <- ggplot(ChickWeight, aes(x = Time, y = weight, colour=Diet)) +
       geom_point() +
  geom_smooth(method =lm)+
facet_wrap(~Diet, nrow=1)


```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
msleep

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x = order))+
  geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x = order))+
  geom_bar()+
  theme(axis.text.x=element_text(angle=90, size=rel(1.5), hjust=1, vjust=0.4))


```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order))+
  geom_bar()+
  theme(
    axis.text.x=element_text(angle=90, size=rel(1.5), hjust=1, vjust=0.4),
    axis.title=element_text(size=rel(1.5), )
    ) +
  labs(x= "Order", y="Observations")





```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill=genus)) +
  theme(
    axis.text.x=element_text(angle=90, size=rel(0.9), hjust=1, vjust=0.4),
    axis.title=element_text(size=rel(1.5))
    legend.key.size=element_rect(size=)
    ) +
  labs(x= "Order", y="Observations")
  





```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill=genus)) +
  theme(
    axis.text.x=element_text(angle=90, size=rel(1.1), hjust=1, vjust=0.4),
    axis.title=element_text(size=rel(1.2)),
    legend.key.size= unit (2, "mm"), 
    legend.text= element_text(size=rel(0.6))
    ) +
  labs(x= "Order", y="Observations")






```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill=genus)) +
  theme(
    axis.text.x=element_text(angle=90, size=rel(1.1), hjust=1, vjust=0.4),
    axis.title=element_text(size=rel(1.2)),
    legend.text= element_text(size=rel(0.6))
    ) +
  labs(x= "Order", y="Observations") +
guides(fill = guide_legend(ncol=3))







```

- Add a contour of thin black lines to the bars.

```{r}


ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill=genus), colour="black", size = 0.1) +
  theme(
    axis.text.x=element_text(angle=90, size=rel(0.9), hjust=1, vjust=0.4),
    axis.title=element_text(size=rel(1.2)),
    legend.text= element_text(size=rel(0.5))
    ) +
  labs(x= "Order", y="Observations") +
guides(fill = guide_legend(ncol=3))







```

- Assign that last plot to an object called `obds_msleep`.

```{r}

obds_msleep <- ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill=genus), colour="black", size = 0.35) +
  theme(
    axis.text.x=element_text(angle=45, size=rel(1), hjust=1, vjust=1),
    axis.title=element_text(size=rel(1.2)),
    legend.key.size= unit (2, "mm"),
    legend.text= element_text(size=rel(0.8)),
   
    ) +
  labs(x= "Order", y="Observations", fill="Genus") +
guides(fill = guide_legend(ncol=3))


obds_msleep





```
```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill=genus), colour="black", size = 0.35) +
  theme(
    axis.text.x=element_text(angle=45, size=rel(1), hjust=1, vjust=1, colour="black"),
    axis.text.y=element_text(size=rel(1.5), colour="black"),
    axis.title=element_text(size=rel(1.2)),
    legend.key.size= unit (2, "mm"),
    legend.text= element_text(size=rel(0.8)),
   
    ) +
  labs(x= "Order", y="Observations", fill="Genus") +
guides(fill = guide_legend(ncol=3))

```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=12, fig.width=12}
library(cowplot)



first_row <- plot_grid(obds_diamonds, obds_chickweight, labels = c("A", "B"), ncol=2, nrow=1)
super_plot <- cowplot::plot_grid(first_row, obds_msleep, labels=c("", "C"), nrow=2)


super_plot
```


```{r, fig.height=12, fig.width=12}
first_row <- cowplot::plot_grid(
  
  
)
second_row <- cowplot::plot_grid(
  
)
super_plot <- cowplot::plot_grid(
  
  
)
super_plot
```

Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)

(obds_diamonds + obds_chickweight)/ (obds_msleep)

```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}
library(gridExtra)

gridExtra::grid.arrange(obds_diamonds, obds_chickweight, ncol=2, nrow=1)



```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave(
  "super_plot.pdf",
  super_plot,
  width=10, height=10
)
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}




```

