---
title: "DESeq2 Tutorial"
author: "Charlie George & Modified by Devika Agarwal"
date: "`r Sys.Date()`"
output: html_document
---

Useful resources 

- **Vignette** - http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html
- **Book** - http://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

# install packages 

- If i didn't already have these installed in my environment - I would delete the # on lines 19-23 and 26-39 and run the code 

```{r}
# from cran 
# install.packages('tidyverse')
# install.packages("RColorBrewer")
# install.packages('pheatmap')
# if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

# from bioconductor
# BiocManager::install("ashr")
# BiocManager::install("apeglm")
# BiocManager::install("DESeq2")
# BiocManager::install("org.Mm.eg.db")
```

# load packages

- what do these packages do? Add comments to remind you what they do 
```{r}
library("pheatmap")   
library("RColorBrewer")
library(tidyverse)  # easier to read and type, more friendly with Python   
library(DESeq2)      
library(ashr)          
library(apeglm)      
library(org.Mm.eg.db) # library for mouse ensembl gene database. Ensembl gene database
```

# 1 Prepping our input data to make deseq2dataset object

## Import counts table

- We made this (or a very similar version) using featurecounts in week1 on the cluster
- raw data -> fastqc -> hisat2 -> featurecounts 

```{r}
# Import counts data to variable called featurecounts_table
# The file has a comment line we need to ignore at top of file 
featurecounts_table <- read_tsv("/t1-data/project/obds/shared/resources/3_r_stats_genomics/rnaseq_differential_expression/allsamples_counts.tsv.gz", comment = "#")
head(featurecounts_table)

# lets save the gene info (Chr,Start,End,Strand, Length) columns as a seperate dataframe (df) incase we need it later
# call this table gene_info and set row names as Geneid
gene_info <- featurecounts_table %>%
                column_to_rownames("Geneid") %>%                     
                dplyr::select(Chr, Start, End, Strand, Length)

head(gene_info)

# create a counts table without the gene info (Chr,Start,End,Strand, Length) columns
# to get a clean counts table where:
# rows = genename, columns = samples
counts_table <- featurecounts_table %>%
                  column_to_rownames("Geneid") %>%
                  dplyr::select(!c(Chr, Start, End, Strand, Length)) %>%
                  as.matrix()

head(counts_table)

# check the dimensions of this counts_table table using dim() -> should have 39,179 genes and 8 cols 

dim(counts_table)
```

# Right now lets import and set up our table of sample information

```{r}
sample_info <- read_tsv("/t1-data/project/obds/shared/resources/3_r_stats_genomics/rnaseq_differential_expression/PRJEB18572_sample_info.tsv")
sample_info

# What class is the cell_type currently? 
class(sample_info$cell_type)


# We want to change it to a factor with 2 levels (the hypothesis we will test)
# Which should be our reference level? 

sample_info$cell_type <- factor(sample_info$cell_type, levels = c("CD4", "CD8"))

sample_info
# what other columns might we want to include in our design that we want to change to factors? 
# replicate
sample_info$replicate <- factor(sample_info$replicate, levels = c("1", "2", "3"))
# genotype
sample_info$genotype <- as.factor(sample_info$genotype)


```

# Check the sample names are the same and in the same order in the sample_info and counts_table 

```{r}
# Column names of the count table and row names of the sample info table must be in the same order
# Check this 
 colnames(counts_table) == rownames(sample_info)
rownames(sample_info) <- sample_info$sample_id
sample_info <- sample_info %>% as.data.frame() %>% column_to_rownames("sample_id")
sample_info
# make sample_info row names from sample_id removing the .bam from the counts_table
colnames(counts_table)

# remove .bam from colnames 
# We are comparing counts_table and sample_info, so the sample_id should be the same between them
colnames(counts_table) <- str_replace(colnames(counts_table), pattern = ".bam", "")
counts_table

colnames(counts_table)==rownames(sample_info)

```

# Generate DESeqDataSet object

```{r}
# look up function to use from slides, find help and create an object called dds
# countData = counts_table
# colData = sample_info
# Experimental design = what is being controled for and what the question is
dds <- DESeqDataSetFromMatrix(countData = counts_table, colData = sample_info, design = ~replicate + genotype + cell_type)

dds
```

## Investigate the dds object 

```{r}
# what does it look like? - e.g. use str(dds)
str(dds)

# how to access counts

counts(dds)

# access colData i.e the sample info table - access sample info from table
colData(dds) # access colData slot(sample metadata)

# access rowData - features table

rowData(dds) # access rowData
# access the design
design(dds)

# does it have any rowRanges()?

rowRanges(dds)
# what are its rownames and colnames? - of counts table
rownames(dds)
colnames(dds)
head(dds)
# what attributes does it have?
attributes(dds)

```

## How to subset a DESeqDataSet Object 

```{r}
# practice subsetting dds object
# select first 3 columns

dds[,1:3]

```

# Filter out unexpressed genes 

- It's good practice to filter out genes with 0 counts across all samples at the start 
- This reduces the size of the object and makes computations quicker without affecting the results as these genes are filtered out anyway in the analysis 
- Note that more strict filtering to increase power is automatically applied via independent filtering on the mean of normalized counts within the results function.


```{r}
# filter out genes that have 0 counts across all samples and save as an object called `dds_filtered`

dds_filtered_TorF <- rowSums(counts(dds))>0


dds_filtered <- dds[rowSums(counts(dds))>0, ]

dds_filtered 
# check dims of dds_filtered  
  
  
```

# Perform exploratory analysis on dds_filtered object 

1. transform dds_filtered object using `rlog()` functions
    - name your new variables `rlog_transformed_dds` 
    - we want to set `blind = TRUE` as because we are doing a QC step we do not want the function to take into account the experimental design, we want it to be an unbiased look at how the samples cluster
    - check the class of the rlog_transformed_dds objects 
    - you can look at the tranformed counts table using `assay()` 

NOTE we only use this tranformed data for QC not for downstream analysis or differential expression itself!!


```{r}
#transform data using `rlog()` functions
rlog_transformed_dds <- rlog(dds_filtered, blind = TRUE)

rlog_transformed_dds

# Notes on `blind` parameter: 
  # blind = TRUE is default
  # blind = TRUE is not the appropriate choice if you expect many or majority of genes to have large differences in counts explainable by   experimental design
  # Mike Love recommends blind = FALSE for visualisation, clustering and machine learning but blind = TRUE for QC
  # Experimental design is only used for calculating the gene-wise dispersion estimates in order to fit a trend line through the dispersions over the mean

  
#check the class of these variable objects 
class(rlog_transformed_dds)
  
#get and look at the tranformed counts table using `assay()`
assay(rlog_transformed_dds)

```

# Optional: If you have time do the tranformation again but using the vst() function 

- As a guide 
    - rlog - more robust than VST when size factors vary widely (10x)
    - VST - less sensitive to high count outliers than rlog, quicker - recommended most often 

```{r}
# get vst transformed data using vst() 
vst_transformed_data <- vst(dds_filtered, blind=TRUE)
vst_transformed_data
```


# Now plot heatmap of sample-sample correlations using rlog transformed count data and cor()


```{r}
# get correlation matrix of the rlog_transformed counts matrix to get correlations between samples 

corr_dds_spearman <- cor(assay(rlog_transformed_dds), method="spearman")

corr_dds_pearson <- cor(assay(rlog_transformed_dds), method="pearson")

# use pheatmap to plot the correlation matrix 

pheatmap(corr_dds_pearson)
```

# try ploting a heatmap of sample distances 

```{r}
# get sample distances -> Annotated what this code is doing?
sampleDists <- dist(t(assay(rlog_transformed_dds))) 
sample_distances_matrix <- as.matrix(sampleDists) 

pretty_blue_colors <- colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)

pheatmap(sample_distances_matrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists, 
         color = pretty_blue_colors)
```

What do you notice here!? what does it tell us about our data? 

Answer: 



## You can subset a dds object and interact with it like you would a DataFrame

```{r}
# remove the two suspect columns using [,] notation - remove suspicious samples 
colnames(rlog_transformed_dds)

rlog_transformed_dds_subset <- as.data.frame(assay(rlog_transformed_dds))  %>% dplyr::select(-"ERR1755088", -"ERR1755089")

rlog_transformed_dds_subset2 <- rlog_transformed_dds[,1:6]

colnames(rlog_transformed_dds_subset)
# lets replot the correlation heatmap with the subsetted data - copy code and adjust from above
class(rlog_transformed_dds_subset)

sampleDists <- dist(t(assay(rlog_transformed_dds_subset2))) 
sample_distances_matrix <- as.matrix(sampleDists) 

pretty_blue_colors <- colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)

pheatmap(sample_distances_matrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists, 
         color = pretty_blue_colors)


```

What do you notice now about the data and the clustering?


# Now lets use the plotPCA function from DESeq2 to plot PCA of rlog_transformed_data_subset

```{r, fig.asp=1}
# plot pca of rlog_transformed_dds using plotPCA()

plotPCA(rlog_transformed_dds_subset2, intgroup= "cell_type", ntop= 500)
# what does the intgroup parameter do? have a go at changing it...

plotPCA(rlog_transformed_dds_subset2, intgroup= "cell_type", ntop= 1000)

# how many variable genes is it using for pca? 


# plot the pca for all genes 

```


What does it tell us about our sample?

Answer: 


------------- This is the end of our Exploratory data analysis -----------------

We now go back to our original dds object with the raw counts and can update it with anything we need to from the exploratory analysis 
    e.g. - update design with variables in sample info table we want to regress out 
         - remove samples (poor quaility or those with name switches etc)
         
# Revisiting our inital data 

What do we need to do to our data??? 

## Update dds_filtered object if needed - call it dds_filtered_updated

```{r}

# do we need to remove/edit samples? 
dds_filtered
dds_filtered_updated <- dds_filtered[,1:6] #keeping first six columns
dds_filtered_updated

# do we want to update the design? Is this nessecary in this case?
# e.g. design(dds) <- ~replicate + condition
design(dds_filtered_updated) <- ~replicate + cell_type


```


# Now we can start our actual DE!!!! yay 

Run deseq2 modelling using one function (DESeq())

```{r}

# look at dds slots then run deseq2 function

DESeq(dds)
# what has happened to your dds slots?
str(dds)
# run deseq2 modelling on the updated dds object in one function using DEseq() 
dds_filtered_updated <- DESeq(dds_filtered_updated)

# then extract the results
results_obj <- results(dds_filtered_updated)

# sumarise results 
summary(results_obj)

```

Have a look at the outputs 

# Now we will run the analysis step by step to see whats going on 

## 1. Estimate size factors using estimateSizeFactors()

```{r}
# Calculate the size factors for each sample using estimateSizeFactors()

dds_filtered_updated <- estimateSizeFactors(dds_filtered_updated)

# Access the size factors from the dds_filtered_updated object

sizeFactors(dds_filtered_updated)

# Do you see any systematic differences in size factors between groups?
#
```

# Optional: Plot size factors as barchart using ggplot
```{r}

size_factors <- sizeFactors(dds_filtered_updated)
size_factors <- data.frame(sample = names(size_factors),
                           size_factor = size_factors,
                           sample_group = colData(dds_filtered_updated)$cell_type)

ggplot(size_factors, aes(x = sample, y = size_factor, fill = sample_group)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 45, h = 1))
```

## Get counts table normalised using size factors

```{r}
# get counts table normalised using size factors - call it normalized_counts

normalized_counts <- counts(dds_filtered_updated, normalized=TRUE)
normalized_counts

# check the class of the output
class(normalized_counts)
```

# 3. Estimate dispersions for each gene

```{r}
# Obtain dispersion estimates for each gene using estimateDispersions() 

dds_filtered_updated <- estimateDispersions(dds_filtered_updated)
dds_filtered_updated
# Plot the per-gene dispersion estimates (DESeq2 has a helper function for this)
plotDispEsts(dds_filtered_updated)

# blue circles indicate low number of biological replicates
```


## 4. Perform Wald test

```{r}
# Tests for significance of coefficients in a negative binomial GLM, using previously calculated size factors and dispersion estimates and WaldTest
dds_filtered_updated <- nbinomWaldTest(dds_filtered_updated)


```


# So we have the actual steps that we've gone through to get to DE in one place generate the dds object again and use the DESeq() function to perform steps 1-3 in one go

```{r}
# generate dds object 

dds_actual_analysis <- DESeqDataSetFromMatrix(countData = counts_table, colData = sample_info, design = ~replicate + cell_type)

# filter rows with 0 counts 

dds_actual_analysis <- dds_actual_analysis[rowSums(counts(dds_actual_analysis))>0, ]

# remove unwanted samples 

dds_actual_analysis <- dds_actual_analysis[,1:6] #keeping first six columns

# run DESeq() 

dds_actual_analysis <- DESeq(dds_actual_analysis)
str(dds_actual_analysis)

# Plot the per-gene dispersion estimates 

plotDispEsts(dds_actual_analysis)
```

# Get the results from your dds object - call it `results_obj`

```{r}
# What do the results() and summary() do, what arguments can you specify? 

results_obj <- results(dds_actual_analysis, alpha= 0.01, contrast=c("cell_type", "CD8", "CD4")) #alpha is the false discovery rate, 10% of the results can be false positives, so setting it to 0.01 means 1% of the genes could be false positives
# Pairwise comparison
summary(results_obj)

results_obj <- as.data.frame(results_obj)


```

# Optional - if we have time: Access the coefficients of the Negative Binomnial GLM – see vingette("DESeq2")
```{r}
head(coef(dds_actual_analysis))
coef_dds <- coef(dds_actual_analysis)
head(coef_dds)
# log fold changes that occurs in the other conditions in each group
# Intercept is absolute gene expression in baseline group in rea
# Coefficients are the logFC in the other conditions
# Coefficients are the log2 fold changes for each group
```

# Plot normalised counts for ENSMUSG00000041779
 
 checkout the 'plotCounts()` function
 
```{r}
# Plot normalised counts for gene ENSMUSG00000041779
normalized_counts <- counts(dds_actual_analysis, normalized = TRUE)
gene_counts <- normalized_counts["ENSMUSG00000041779", ] %>%
  as.data.frame() %>%
  rownames_to_column("sample") 
  
  colnames(gene_counts) <- c("sample", "count")
  colData(dds_actual_analysis)$cell_type 
  gene_counts <- gene_counts %>% 
    mutate("cell_type"=colData(dds_actual_analysis)$cell_type)
  gene_counts %>%
  ggplot(aes(x=cell_type, y=count)) +
    geom_boxplot() +
    geom_point()

```

# Access the results table for the comparison between CD8 and CD4 T cells . Youu can also try to CD4 vs CD8 (i.e. switch the reference) - you could also try this for replicate 1 vs replicate 3 as a control 

checkout `resultsNames()`  & `results()`

```{r}
#What the variable is, and the control is the first variable
results_obj_names <- results(dds_actual_analysis, contrast=c("cell_type", "CD8", "CD4"), alpha= 0.01)

results_obj_replicate <- results(dds_actual_analysis, contrast=c("replicate", "1", "2"), alpha= 0.01)

head(results_obj_names)

head(results_obj_replicate)
```

Table should contain log2 fold changes and p-values for all genes

# Plot a histogram of the raw and BH-adjusted p-values using ggplot – do they look as expected? for CD8 vs CD4

```{r}
results_obj %>% 
    as.data.frame() %>% 
    ggplot() +
    geom_histogram(aes(x=pvalue), fill='lightblue') + 
    geom_histogram(aes(x=padj),fill='lightgreen')

head(results_obj)
class(results_obj)
```


# Generate an MA plot of the log2 FC values for all genes using DESeq2 plotMA()

```{r}
plotMA(results_obj_names, alpha=0.01, ylim=c(-10,10))

class(results_obj_names)
```

# Shrink the log2 FC values using the normal, apeglm and ashr methods -> see the vignette – which shall we choose? 

```{r}

# call your results table  called shrunkenLFC_dds_results_obj
resultsNames(dds_actual_analysis)

shrunken_dds_results_apeglm <- lfcShrink(dds_actual_analysis, coef="cell_type_CD8_vs_CD4", type="apeglm")

plotMA(shrunken_dds_results_apeglm, alpha=0.01, ylim=c(-10,10))

plotMA(results_obj_names, alpha=0.01, ylim=c(-10,10))

lfcShrink(dds_actual_analysis, coef = "cell_type_CD8_vs_CD4", type="ashr") %>% 
    as.data.frame() %>% 
    ggplot() +
    geom_point(aes(x=log2(baseMean),y=log2FoldChange,color=padj < 0.05),size=0.5) 
```


# Generate a results table (one shrinkage method) containing gene symbols and (if time) entrez ids 

```{r}

# convert shrunkenLFC_dds_results_obj to a dataframe and call it shrunkenLFC_dds_results_df

head(shrunken_dds_results_apeglm)

# Use the org.Mm.eg.db package (hint:you did this yesterday)
# Remove all genes with a padj of NA
# How many Ensembl IDs are not assigned an mgi symbol and how many mgi symbols are duplicated?

keytypes(org.Mm.eg.db)      #to remind us of what keytypes are in this particular db

# # mapIds gives you a 1:1 mapping by default (can only extract one column at a time), whereas the select function doesn't (but you can extract multiple columns at a time)

MGI_symbols <- mapIds(org.Mm.eg.db, row.names(shrunken_dds_results_apeglm), "MGI", "ENSEMBL") #indicating the row names from that database. Getting out the anotations "MGI and ENSEMBL"
gene_symbols <- mapIds(org.Mm.eg.db, row.names(shrunken_dds_results_apeglm), "SYMBOL", "ENSEMBL")
entrez_ids <- mapIds(org.Mm.eg.db, row.names(shrunken_dds_results_apeglm), "ENTREZID", "ENSEMBL")

table(names(entrez_ids)==row.names(shrunken_dds_results_apeglm))     #to check that the order of the vectors are the same as the df
#options(scipen=0)

#add mapId vectors to the original results df
results_shrunken_LFC_apeglm_filtered_df <- shrunken_dds_results_apeglm %>%
    as.data.frame() %>% 
    mutate(mgi_symbols=MGI_symbols) %>%
    mutate(gene_symbols=gene_symbols) %>%
    mutate(ENTREZ_IDs=entrez_ids)


# How many Ensembl IDs are not assigned an mgi symbol and how many mgi symbols are duplicated?
which(duplicated(MGI_symbols)) # looking how many duplicates are in the symbols
which(is.na(MGI_symbols)) #which genes have NA in the MGI_symbol
```


# Export all the results to CSV

```{r}
write_csv(results_shrunken_LFC_apeglm_filtered_df, "/project/obds/vpereira/DESeq2_tutorial/CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv")

# to specify the path

# write_csv(results_shrunken_LFC_apeglm_filtered_df, "/project/obds/vpereira/DESeq2_tutorial/CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv")
```

# Filter res_df and save as CSV

```{r}

results_shrunken_LFC_apeglm_filtered_df <- results_shrunken_LFC_apeglm_filtered_df %>% 
  rownames_to_column("ensembl_gene_id")

write_csv(results_shrunken_LFC_apeglm_filtered_df, "/project/obds/vpereira/DESeq2_tutorial/CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv")

#Filter the results table for padj < 0.05 and absolute log2 FC > 1, and write to a CSV file
res_0.05 <- results_shrunken_LFC_apeglm_filtered_df %>% 
    dplyr::filter(padj < 0.05 & abs(log2FoldChange) > 1)

#Get upregulated genes - Filter the results table for padj < 0.05 and log2FoldChange FC > 1, and write to a CSV file
res_0.05_up <- results_shrunken_LFC_apeglm_filtered_df %>% 
    dplyr::filter(padj < 0.05 & log2FoldChange > 1)

#Get doownregulated genes - Filter the results table for padj < 0.05 and log2FoldChange FC < 1, and write to a CSV file
res_0.05_down <- results_shrunken_LFC_apeglm_filtered_df %>% 
    dplyr::filter(padj < 0.05 & log2FoldChange < -1)

write_csv(res_0.05, 
          "CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1.csv")

write_csv(res_0.05_up,
          "CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_up.csv")

write_csv(res_0.05_down,
          "CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_down.csv")
```


Well done! That is the basics of exploratory data analysis for bulk RNAseq and differential gene expression - these statistical methods underly the foundations for single cell differential expression analysis, although scRNA-seq has a few extra paramterisations and complications to take into account. 

We will now progress to pathway/geneset analysis to get some biological meaning from our DE gene list....

