---
title: "20221021_Pathway analysis"
author: "Vinnycius P Almeida"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

This is the pathway analysis practical of the OBDS course 
- The first part of this pracical is using gprofiler to perform ORA on output dataset of differentially expressed genes from DESeq2

# load packages 
```{r}
# load librarys

library(tidyverse)
library(gprofiler2)
```

# Read in our tables from this morning in a new Rmd file
```{r}
# read in files from this morning 

# 1. all results (this was filtered to remove non DE genes in our DESeq2 list so only contains genes expressed in our samples) -> call this table 'all_genes_df'  - this will be our background
all_genes_df <- read_csv("/project/obds/vpereira/DESeq2_tutorial/CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv")

#filter out the genes which padj is NA 
all_genes_df_filtered <- all_genes_df %>% 
    dplyr::filter(!is.na (padj))

# 2. all significant upregulated genes -> call this upreg_genes_df
upreg_genes_df <- read_csv("/project/obds/vpereira/DESeq2_tutorial/CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_up.csv")

# 3. all significant down regulated genes -> call this downreg_genes_df 
downreg_genes_df <- read_csv("/project/obds/vpereira/DESeq2_tutorial/CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_down.csv")



# Check the dimensions of each of your dataframes using dim()
dim(all_genes_df)
dim(all_genes_df_filtered)
dim(upreg_genes_df)
dim(downreg_genes_df)
# upreg_genes_df should be 915
# downreg_genes_df should be 1204
# all_genes_df_filtered should be 17729


```
Now we want a vector of gene ids to use for analysis 

```{r}
# from each of these dataframes get a vector of the ensembl ids 
all_gene_ids <- all_genes_df_filtered$ensembl_gene_id
    
upreg_gene_ids <- upreg_genes_df$ensembl_gene_id
    
downreg_gene_ids <- downreg_genes_df$ensembl_gene_id
```

#Perform ORA using g:profiler 

1. Look at documentation - https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
```{r}
# use the gost() function - what paramters should we include/change 
# we want to test the ORA of upreg_gene_ids vs our all_gene_ids

# get help for function
?gost() 

# write query and assign to output variable named `gost_results_obj`

gost_results_obj <- gost(query = upreg_gene_ids, 
                organism = "mmusculus", ordered_query = FALSE, 
                multi_query = FALSE, significant = TRUE, exclude_iea = TRUE, 
                measure_underrepresentation = FALSE, evcodes = FALSE, 
                user_threshold = 0.01, correction_method = "g_SCS", 
                domain_scope = "annotated", custom_bg = all_gene_ids, 
                numeric_ns = "", sources = NULL, as_short_link = FALSE)

                gost(query = upreg_gene_ids, 
                organism = "mmusculus", ordered_query = FALSE, 
                multi_query = FALSE, significant = TRUE, exclude_iea = TRUE, 
                measure_underrepresentation = FALSE, evcodes = FALSE, 
                user_threshold = 0.01, correction_method = "g_SCS", 
                domain_scope = "annotated", custom_bg = all_gene_ids, 
                numeric_ns = "", sources = NULL, as_short_link = TRUE)

# multi-query = multiple list of genes
# custom_bg = baseline genes
# what are our outputs  
str(gost_results_obj)

```

```{r}
# plot our results using in built function in gprofiler -> which one is it? 

gostplot(gost_results_obj, capped = TRUE, interactive = TRUE) 
```

```{r}
# what does as_short_link = FALSE do in gost() function? 

# indicator to return results as short-link to the g:Profiler web tool. If set to TRUE, then the function returns the results URL as a character string instead of the data.frame. Creates an interactive link
```

Answer -> 

```{r}
# Filter results table for genesets with between 5-300 genes 

# done in the website https://biit.cs.ut.ee/gprofiler/gost
```


2. Try querying multiple gene lists at the same time (up and down regulated genes) using Multiple queries - see the documentation 

```{r}

```

Can you plot these querys together?

```{r}


```

Output the get_version_info() to save the versions of the databases that were used for the analysis!! 
```{r}
get_version_info(organism = "mmusculus")

```


```{r}
sessionInfo()

```