---
title: Week4-day5 :integration template"
author: "Devika Agarwal"
date: "21/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #print code and results
```


```{r}
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
library(patchwork)
library(tidyverse)
library(batchelor)
library(harmony)
library(lisi)
library(MAST)
options(future.globals.maxSize = 4000 * 1024^2)    # Maximum allowed total size (in bytes) of global variables identified
```

# Exercise 1 - joint analysis of two PBMC datasets

Here we will look at how to PBMC datasets compare without any integration. 

## 1. Import the two RDS files into R

These RDS object have already been put through the QC and filtering steps covered on Seurat day 1.


```{r}
pbmcv2 <- readRDS("pbmcv2_filtered.rds")
pbmcv3 <- readRDS("pbmcv3_filtered.rds")
```


Plot these metadata covariates nCount_RNA/nFeature_RNA/percent.mt to see how they range for both the datasets 

use `VlnPlot()` from Seurat

```{r}

head(pbmcv2)

p1 <- VlnPlot(object= pbmcv2, features= c("nCount_RNA", "nFeature_RNA", "percent.mt"))
p2 <- VlnPlot(object=pbmcv3, features= c("nCount_RNA", "nFeature_RNA", "percent.mt"))

p1 / p2
```



## 2. Create a merged Seurat object containing both sets of data (no batch correction)

Vignette: https://satijalab.org/seurat/articles/merge_vignette.html

```{r}

pbmcv2[[]]
pbmcv3[[]]
seurat_merged <- merge(x = pbmcv2, y = pbmcv3, add.cell.ids= c("pbmcv2", "pbmcv3"))




```

Check the dims of the merged object in comparison to the separate objects

What is the total gene number in the merged object compared to each single object?

```{r}
dim(pbmcv2)
dim(pbmcv3)
dim(seurat_merged)

GetAssayData(seurat_merged, "counts")[1:5, 1:5]

```

## 3.1 Perform SCTransform normalisation

Do you want to regress any meta.data variables ?

```{r, warning=FALSE}
?SCTransform
Assays(seurat_merged)
seurat_merged <- SCTransform(object= seurat_merged, assay= "RNA", vars.to.regress= "percent.mt")

```

Plot to see how the nCount_RNA looks before and after SCT normalisation for the merged dataset?

Did SCT do a good job in normalizing the sequencing depth across the two datasets?


```{r}

p_merged <- VlnPlot(object=seurat_merged, features= c("nCount_RNA", "nCount_SCT"), same.y.lims = TRUE) +
  scale_y_continuous(breaks=c(3000, 6000, 9000, 12000))

p_merged
```


What should the Default Assay be?

```{r}
DefaultAssay(seurat_merged)
```

Plot the Variable features plot for data with top 10 variable genes labelled

```{r, fig.width=10}

Assays(seurat_merged)
?VariableFeaturePlot

?VariableFeatures

top10genes <- Seurat::VariableFeatures(object= seurat_merged)[1:10]

p_var <- Seurat::VariableFeaturePlot(object= seurat_merged, selection.method = "sct")

?LabelPoints

p_var_lab <- LabelPoints(p_var, points= top10genes, repel=TRUE, size = 5, label.r= 5); p_var_lab



```

## 3.2 Run PCA and visualize 

Can you see a batch effect between pbmc v2 and v3?


```{r}
?RunPCA
seurat_merged <- RunPCA(object = seurat_merged)
```

Look at which genes are driving your PCs.

```{r, fig.width=8}
?VizDimLoadings
VizDimLoadings(object = seurat_merged, dims= 1)

```


```{r}
?DimPlot
DimPlot(object = seurat_merged, dims= c(1,2))

```

Choose optimal  number of PCs for UMAP

```{r}
?ElbowPlot
ElbowPlot(object= seurat_merged, ndims = 20, reduction = "pca")
```

## 3.3 Run the Seurat UMAP function and visualize

Is there a batch effect? 


```{r}
?RunUMAP
seurat_merged <- RunUMAP(object= seurat_merged, dims=1:20)
?DimPlot
DimPlot(object= seurat_merged, dims=c(1,2))
```

### Optional: Do cells cluster by celltypes?

Plot some pbmc celltype markers to see how they look on the UMAP

```{r, fig.width=10}
?FeaturePlot
p1 <- FeaturePlot(seurat_merged, features = top10genes) 

p2 <- FeaturePlot(seurat_merged, features = c("CD3E", "CD4", "CD8A", "MS4A1", "CD14", "FCGR3A","LYZ","GNLY"))

p2
```


## Exercise 2: Seurat integration with SCTransform

Check out the Vignette: https://satijalab.org/seurat/articles/integration_introduction.html#performing-integration-on-datasets-normalized-with-sctransform-1 

The Seurat v2 integration method is slightly different depending on whether you perform log normalistaion or SCTransformation on your seurat objects. Here we are using the SCTransform method (the vignette covers both).

### 1. Set up

Import the RDS files 

We will read in the objects again to start the integration workflow from scratch

```{r}

pbmcv2 <- readRDS("pbmcv2_filtered.rds")
pbmcv3 <- readRDS("pbmcv3_filtered.rds")

```

Generate a named list of the two objects to run workflow

```{r}
pbmc.list <- list("v2"= pbmcv2, "v3" = pbmcv3)
```

Apply SCTransform across the list

```{r}
?SCTransform
pbmc.list <- lapply(pbmc.list, FUN = SCTransform, vars.to.regress="percent.mt") 
```

check dims for both objects in list after SCTransform
```{r}
dim(pbmc.list$v2) #or dim(pbmc.list[[v2]])

dim(pbmc.list[["v3"]])
```

make sure Default assay in both objects is SCT
```{r}
DefaultAssay(pbmc.list$v2)

DefaultAssay(pbmc.list$v3)
```

Choose the features to use when integrating multiple datasets. 

This function ranks features by the number of datasets they are deemed variable in, breaking ties by the median variable feature rank across datasets. It returns the top scoring features by this ranking.
```{r}
?SelectIntegrationFeatures
features <- SelectIntegrationFeatures(object.list= pbmc.list, nfeatures = 3000) 
```

Prepare an object list normalized with sctransform for integration (run `PrepSCTIntegration()` )

This function subsets the scale.data slot to only contain residuals for anchor.features for efficiency in downstream processing
It is specific to the SCTransform workflow and makes sure you have the SCTransform residuals for the remaining genes

```{r}
?PrepSCTIntegration
pbmc.list <- PrepSCTIntegration(object.list= pbmc.list, anchor.features = features)

```

> After this step, the features present in the scale.data slots of the two objects are the same
> anchor.features are used in the anchor finding process


### Optional exercise

How similar are these features to the variable features that we found for the pbmcv2 and v3 individually?
- extract variable features for each dataset individually
- compare these with "features"; which has the variable features most common for both datasets


## 2. Perform integration

Find a set of anchors between a list of Seurat objects. These anchors can later be used to integrate the objects.

```{r}
?FindIntegrationAnchors
pbmc.anchors <- FindIntegrationAnchors(object.list= pbmc.list, anchor.features = features, normalization.method= "SCT", dims=1:20, reduction = "cca")

```

Integrate the two pbmc objects

```{r}
?IntegrateData
pbmc.combined.sct <- IntegrateData(anchorset= pbmc.anchors, normalization.method= "SCT", dims=1:20)

class(pbmc.anchors)
```

> If normalization.method = "SCT", the integrated data is returned to the scale.data slot and can be treated as centered, corrected Pearson residuals. so we dont scale again after integrating the two datasets


what is the Default assay now?
Check to see if the seurat object has changed in the Assay slot at all

```{r}
DefaultAssay(pbmc.combined.sct)
```

## 3. Analysis of integrated data

Run  PCA on Integrated data

```{r}
?RunPCA
pbmc.combined.sct <- RunPCA(object= pbmc.combined.sct)
```

How many PCs should we use going forward?
```{r}
ElbowPlot(pbmc.combined.sct)
```

Visualise PCA, how does this compare to the merged data?

```{r}

?DimPlot
p1 <- DimPlot(object=pbmc.combined.sct, dims=c(1,2)) + ggtitle("Integrated")

p2 <- DimPlot(object=seurat_merged, dims=c(1,2), reduction= "pca") + ggtitle("Merged")

p1 + p2
```



Run UMAP 
```{r}
?RunUMAP
pbmc.combined.sct <- RunUMAP(object= pbmc.combined.sct, dims=1:20)

p1 <- DimPlot(object=pbmc.combined.sct, dims=c(1,2), reduction = "umap") + ggtitle("Integrated")

p2 <- DimPlot(object=seurat_merged, dims=c(1,2), reduction = "umap") + ggtitle("Merged")

p1 + p2
```


```{r}
DefaultAssay(pbmc.combined.sct)

Assays(pbmc.combined.sct)
```

## 4. Identify markers

We want to identify markers that distinguish each cluster. 
Look up `FindMarkers` and `FindAllMarkers`, which one should we use here?


> Answer

Why are we changing the DefaultAssay?
Do we need to renormalise RNA Assay?

> We can't run FindMarkers on the integrated SCT model because the genes are no longer independent of each other. 
> Therefore it does not meet the assumptions that are required for DE. 

> NB if using SCTransform v2 there is an additional step ` PrepSCTFindMarkers` and then FndAllMarkers can be run on the SCT assay
> Here we will run it on the RNA assay as usual


```{r}

DefaultAssay(pbmc.combined.sct) <- "integrated"
DefaultAssay(pbmc.combined.sct)

pbmc.combined.sct <- FindNeighbors(pbmc.combined.sct, 
                                   reduction= "pca", 
                                   dims=1:20) # findneighbors have been run in "integrated"

Graphs(pbmc.combined.sct) 

?FindClusters

pbmc.combined.sct <- FindClusters(object= pbmc.combined.sct, resolution= c(0.8, 0.5)) #adding the resolution columns to the dataframe


head(pbmc.combined.sct)
pbmc.combined.sct$seurat_clusters[1:5]
```

How many clusters have been defined at each resolution?

```{r}
# 8
```


Visualse UMAP plot for batch effect and clusters

check the Default Assay to be sure

```{r}


DimPlot(object=pbmc.combined.sct, dims=c(1,2), reduction = "umap", group.by= c("integrated_snn_res.0.8", "integrated_snn_res.0.5")) 


```

Plot some canonical PBMC cluster markers

Which assay should we plot them on? (try both Integrated and SCT, is there a difference?)


```{r, fig.width=8}
DefaultAssay(pbmc.combined.sct) <- "integrated"
p1 <- FeaturePlot(object = pbmc.combined.sct, features = c("S100A9", "S100A8", "LYZ", "IGKC","GNLY","IGHM", "NKG7", "IGLC2")) 


DefaultAssay(pbmc.combined.sct) <- "SCT"
p2 <- FeaturePlot(object = pbmc.combined.sct, features = c("S100A9", "S100A8", "LYZ", "IGKC","GNLY","IGHM", "NKG7", "IGLC2")) 
 # SCT assay is better to visualise the gene expression in different clusters. It is "cleaner"
p1 
p2 
```

## 4. Identify markers

Differential expression for cluster markers
look up `FindMarkers` and `FindAllMarkers`

Why are we changing the DefaultAssay?
Do we need to renormalise RNA Assay?

Use the "LR" test

```{r}
DefaultAssay(pbmc.combined.sct) <- "RNA"

head(pbmc.combined.sct)
?FindMarkers
pbmc.combined.sct.markers.LR <- FindAllMarkers(
  object= pbmc.combined.sct, 
  logfc.threshold = 0.25,
  test.use = "LR",
  min.pct = 0.1,
  latent.vars = "orig.ident")
    
```

Filter the top 5 markers for each cluster from the LR test
use `dply::group_by` and `slice_min()`
How do you define "top"?
```{r}
?group_by

pbmc.combined.sct.markers.top5 <- pbmc.combined.sct.markers.LR %>%
  dplyr::group_by(cluster) %>%
  slice_min(n=5, order_by= p_val_adj); pbmc.combined.sct.markers.top5 

```

plot Heatmap or Dotplot of markers

try functions `Seurat::DoHeatmap` or `Seurat::Dotplot`

```{r}

DefaultAssay(pbmc.combined.sct) <- "SCT"
?Seurat::DoHeatmap

Seurat::DoHeatmap(object=pbmc.combined.sct, features= pbmc.combined.sct.markers.top5$gene, assay= "SCT") + ggtitle("Top5")


```

```{r}
?Seurat::DotPlot
Seurat::DotPlot(object=pbmc.combined.sct, features= unique(pbmc.combined.sct.markers.top5$gene), assay= "SCT") +RotatedAxis() + ggtitle("Top5")
```
top3
```{r}
DefaultAssay(pbmc.combined.sct) <- "SCT"
?Seurat::DoHeatmap

pbmc.combined.sct.markers.top3 <- pbmc.combined.sct.markers.LR %>%
  dplyr::group_by(cluster) %>%
  slice_min(n=3, order_by= p_val_adj)

Seurat::DoHeatmap(object=pbmc.combined.sct, features= pbmc.combined.sct.markers.top3$gene, assay= "SCT") + ggtitle("Top3")

```
```{r}
?Seurat::DotPlot
Seurat::DotPlot(object=pbmc.combined.sct, features= unique(pbmc.combined.sct.markers.top3$gene), assay= "SCT") +RotatedAxis() + ggtitle("Top3")
```



How well do you think the clustering has worked?
Are you able to identify each cluster?
Would you re-evaluate any of the clusters?


## Optional 

Spend some time experimenting with different statistical tests in the FindAllMarkers command
Try some different parameters
Sort by log2fc vs padj
How does this affect the plots?
Use your results to label the identity of each cluster and plot on a UMAP

```{r}
library(Seurat)
library(cowplot)
library(harmony)

harmony.onehot.new <- function (x)
{
  data.frame(x) %>% tibble::rowid_to_column("row_id") %>% dplyr::mutate(dummy = 1) %>%
    tidyr::spread(x, .data$dummy, fill = 0) %>% dplyr::select(-row_id) %>%
    as.matrix
}
environment(harmony.onehot.new) <- asNamespace('harmony')
assignInNamespace("onehot", harmony.onehot.new, ns = "harmony")

Reductions(seurat_merged)
?RunHarmony
table(seurat_merged$orig.ident)
```

```{r}
options(repr.plot.height = 2.5, repr.plot.width = 6)
seurat_merged_harmony <- seurat_merged %>% 
    RunHarmony("orig.ident", plot_convergence = TRUE, assay.use= "SCT", dims.use=1:30)


```

#Bonus Exercise: Run Harmony based integration

You can run Harmony within your Seurat workflow. You'll only need to make two changes to your code.

IMPORTANT DIFFERENCE: In the Seurat integration tutorial, you need to define a Seurat object for each dataset. 
With Harmony integration, create only one Seurat object with all cells.
Before running Harmony, make a Seurat object and following the standard pipeline through PCA.

After that Run Harmony with the `RunHarmony()` function
In downstream analyses, use the Harmony embeddings instead of PCA.
For example, run Harmony and then UMAP in two lines.

Vignette: http://htmlpreview.github.io/?https://github.com/immunogenomics/harmony/blob/master/docs/SeuratV3.html


Create a merged object, run SCTransform and PCA
```{r}
# create a merged obect of the datasets you want ot integrate
seurat_merged <- merge()

# Normalise the data NormalizeData(), ScaleData() , FindVariablegenes() or SCTransform()

# Run PCA
```


Run Harmony
```{r}

```


Run UMAP, FindNeighbours and FindClusters, then visualise some key genes. 
```{r}
seurat_merged_harmony <- seurat_merged_harmony %>%
  RunUMAP(reduction ="harmony", dims=1:20) %>%
  FindNeighbors(reduction="harmony", dims=1:20) %>%
  FindClusters(resolution=0.5) %>%
  identity()

options(repr.plot.height = 4, repr.plot.width = 10)
DimPlot(seurat_merged_harmony, reduction= "umap", group.by = "orig.ident", pt.size=.1)

options(repr.plot.height = 4, repr.plot.width = 10)
DimPlot(seurat_merged_harmony, reduction= "umap", group.by = "orig.ident", pt.size=.1)
```





